---
title: "Filtering of ChIP-nexus peaks for CSL proteins and Mga2 - using peaks called from pooled replicate data"
author: "Martin Převorovský, prevorov@natur.cuni.cz, http://www.GenoMik.cz"
output: html_document
---

```{r setup, eval=TRUE}
library(GenomicRanges)
library(rtracklayer)

summit_dir <- './peaks_pooled/'
summit_file_extension <- '.bed$'
peaks_filtered_dir <- './peaks_filtered_pooled/'

peak_window <- 50
controls <- c('EXP_Fkh2', 'STAT_Fkh2', 'EXP_untagged', 'STAT_untagged')

dir.create(peaks_filtered_dir, showWarnings = FALSE)
```


* remove peaks in *S. cerevisiae* chromosomes

```{bash, eval=TRUE}
sed -i "/tpg\|ref/d" ./peaks_pooled/*.bed
```


* load BED files with summit info

* extend summits into peaks (using an arbitrary peak window size)

* extract sample info, and list of experimental conditions

```{r, eval=TRUE}
summit_files <- list.files(summit_dir, pattern = summit_file_extension, full.names = TRUE)
samples <- gsub("\\..*", '', basename(summit_files))
samples <- gsub("_summits", '', samples)
conditions <- unique(gsub('^.*?_', '', samples))

peaks <- NULL
for (i in 1:length(summit_files)) {
  current_file <- import.bed(summit_files[i])
  
  current_file$file_name <- summit_files[i]
  current_file$sample <- samples[i]
  current_file$summit <- start(current_file)
  start(current_file) <- start(current_file) - peak_window + 1
  end(current_file) <- end(current_file) + peak_window
  
  peaks <- c(peaks, current_file)
}

rm(current_file)
```

* filter out peaks present in untagged controls

```{r, eval=TRUE}
control_indices <- which(samples %in% controls)
conditions_exp_indices <- grep('Cbf12', samples, invert = TRUE)
conditions_exp_indices <- conditions_exp_indices[!conditions_exp_indices %in% control_indices]
conditions_stat_indices <- grep('Cbf12', samples)

untagged_exp_index <- grep('EXP_untagged', samples)
untagged_stat_index <-grep('STAT_untagged', samples) 
  
peaks_kept <- peaks
peaks_dropped <- peaks

for (i in conditions_exp_indices){
  peaks_kept[[i]] <- subsetByOverlaps(peaks[[i]], peaks[[untagged_exp_index]], invert = TRUE)
  peaks_dropped[[i]] <- subsetByOverlaps(peaks[[i]], peaks[[untagged_exp_index]])
}

for (i in conditions_stat_indices){
  peaks_kept[[i]] <- subsetByOverlaps(peaks[[i]], peaks[[untagged_stat_index]], invert = TRUE)
  peaks_dropped[[i]] <- subsetByOverlaps(peaks[[i]], peaks[[untagged_stat_index]])
}
```

* save filtered peaks as BED files (kept and dropped as separate files)

```{r, eval=TRUE}
sample_indices_to_process <- c(conditions_exp_indices, conditions_stat_indices)
for (i in sample_indices_to_process) {
  filename_peaks <- paste0(peaks_filtered_dir, 
                           samples[i], 
                           '.untagged-filtered.peaks.bed')
  filename_summits <- paste0(peaks_filtered_dir, 
                             samples[i], 
                             '.untagged-filtered.summits.bed')
  
  tmp_bed <- peaks_kept[[i]]
  
  export.bed(tmp_bed, filename_peaks, format = 'bed')
  
  start(tmp_bed) <- tmp_bed$summit
  end(tmp_bed) <- tmp_bed$summit # there is automatic conversion to 0-based [ ) intervals (ie, closed/open)
  export.bed(tmp_bed, filename_summits, format = 'bed')
}

for (i in sample_indices_to_process) {
  filename_peaks <- paste0(peaks_filtered_dir, 
                           samples[i], 
                           '.untagged-filtered_dropped.peaks.bed')
  filename_summits <- paste0(peaks_filtered_dir, 
                             samples[i], 
                             '.untagged-filtered_dropped.summits.bed')
  
  tmp_bed <- peaks_dropped[[i]]
  
  export.bed(tmp_bed, filename_peaks, format = 'bed')
  
  start(tmp_bed) <- tmp_bed$summit
  end(tmp_bed) <- tmp_bed$summit # there is automatic conversion to 0-based [ ) intervals (ie, closed/open)
  export.bed(tmp_bed, filename_summits, format = 'bed')
}

rm(filename_peaks, filename_summits, tmp_bed)
```

* filter out peaks present in Fkh2-TAP (irrelevant IP) controls

```{r, eval=TRUE}
fkh2_exp_index <- grep('EXP_Fkh2', samples)
fkh2_stat_index <-grep('STAT_Fkh2', samples) 
  
peaks_kept_kept <- peaks_kept
peaks_kept_dropped <- peaks_kept

for (i in conditions_exp_indices){
  peaks_kept_kept[[i]] <- subsetByOverlaps(peaks_kept[[i]], 
                                           peaks[[fkh2_exp_index]], invert = TRUE)
  peaks_kept_dropped[[i]] <- subsetByOverlaps(peaks_kept[[i]], 
                                              peaks[[fkh2_exp_index]])
}

for (i in conditions_stat_indices){
  peaks_kept_kept[[i]] <- subsetByOverlaps(peaks_kept[[i]], 
                                           peaks[[fkh2_stat_index]], invert = TRUE)
  peaks_kept_dropped[[i]] <- subsetByOverlaps(peaks_kept[[i]], 
                                              peaks[[fkh2_stat_index]])
}
```

* save filtered peaks as BED files (kept and dropped as separate files)

```{r, eval=TRUE}
for (i in sample_indices_to_process) {
  filename_peaks <- paste0(peaks_filtered_dir, 
                           samples[i], 
                           '.untagged+Fkh2-filtered.peaks.bed')
  filename_summits <- paste0(peaks_filtered_dir, 
                             samples[i], 
                             '.untagged+Fkh2-filtered.summits.bed')
  
  tmp_bed <- peaks_kept_kept[[i]]
  
  export.bed(tmp_bed, filename_peaks, format = 'bed')
  
  start(tmp_bed) <- tmp_bed$summit
  end(tmp_bed) <- tmp_bed$summit # there is automatic conversion to 0-based [ ) intervals (ie, closed/open)
  export.bed(tmp_bed, filename_summits, format = 'bed')
}

for (i in sample_indices_to_process) {
  filename_peaks <- paste0(peaks_filtered_dir, 
                           samples[i], 
                           '.untagged+Fkh2-filtered_dropped.peaks.bed')
  filename_summits <- paste0(peaks_filtered_dir, 
                             samples[i], 
                             '.untagged+Fkh2-filtered_dropped.summits.bed')
  
  tmp_bed <- peaks_kept_dropped[[i]]
  
  export.bed(tmp_bed, filename_peaks, format = 'bed')
  
  start(tmp_bed) <- tmp_bed$summit
  end(tmp_bed) <- tmp_bed$summit # there is automatic conversion to 0-based [ ) intervals (ie, closed/open)
  export.bed(tmp_bed, filename_summits, format = 'bed')
}

rm(filename_peaks, filename_summits, tmp_bed)
```




```{bash, eval=TRUE}
computeMatrix reference-point -S ./coverage_ChIP-nexus/Cbf11*.bw -R ./peaks_filtered_pooled/Cbf11.untagged*summits.bed -a 300 -b 300 -o matrix11_pooled
plotHeatmap -m matrix11_pooled -o test11_pooled.svg --sortUsing max --sortUsingSamples 6

computeMatrix reference-point -S ./coverage_ChIP-nexus/Cbf12*.bw -R ./peaks_filtered_pooled/Cbf12.untagged*summits.bed -a 300 -b 300 -o matrix12_pooled
plotHeatmap -m matrix12_pooled -o test12_pooled.svg --sortUsing max --sortUsingSamples 4
```

```{bash, eval=TRUE}
computeMatrix reference-point -S ./coverage_ChIP-nexus/Cbf11*plus*.bw ./coverage_ChIP-nexus/Mga2*plus*.bw -R ./peaks_filtered_pooled/Cbf11.untagged+Fkh2-filtered.summits.bed -a 400 -b 400 -o matrix_fin11_pooled
plotHeatmap -m matrix_fin11_pooled -o test_fin11_pooled.svg --sortUsing sum --sortUsingSamples 3 --missingDataColor red --colorMap Greys

computeMatrix reference-point -S ./coverage_ChIP-nexus/Cbf12*plus*.bw -R ./peaks_filtered_pooled/Cbf12.untagged+Fkh2-filtered.summits.bed -a 400 -b 400 -o matrix_fin12_pooled
plotHeatmap -m matrix_fin12_pooled -o test_fin12_pooled.svg --sortUsing sum --sortUsingSamples 2 --missingDataColor red --colorMap Greys 
```


```{r}
sessionInfo()
```